---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# rdemtools

The goal of *rdemtools* is to provides functions to edit digital elevation models (**DEMs**). This package is very simple, with only a few functions and without new classes. However, it offers an easy way to use the power of R to edit DEMs.

## Installation

You can install rdemtools from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("GastonMauroDiaz/rdemtools")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r include=FALSE}
require(rdemtools)

```


```{r example}
set.seed(11)
reference <- fake_dem(n_random_data = 60, z_range = c(0, 600))
plot(reference)

set.seed(18)
fillSource <- fake_dem(z_range = c(0, 20))
fillSource <- fillSource - mean(fillSource[])
fillSource <- fillSource + reference
fillSource <- smooth_dem(fillSource, theta = 6)
plot(fillSource)

p <- sampleRegular(reference, 10, sp = TRUE)
voidsMask <- fake_voids(p, void_size = 20, reference)

dem <- reference
dem[voidsMask] <- NA
plot(dem)

slp <- terrain(fillSource, "slope")
asp <- terrain(fillSource, "aspect")
demF2 <- fill_voids_k08(dem, slp, asp, show_progress_bar = FALSE)

plot(demF2)

plot(demF2 - reference)

```

## Vignette

After install *rdemtool*, you can see the vignettes with: 

```{r eval=FALSE}
browseVignettes(package = "rdemtools")
```

